\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{hyperref}

\title{Final Project Report Addendum: Label Handling and Ignore Logic}
\author{Semantic Segmentation Project}
\date{\today}

\begin{document}
\maketitle

\section*{Goal of Label Conversion}
The model is trained with \texttt{NUM\_CLASSES = 20}, so valid trainable labels are only in the range \texttt{0..19}. Raw Cityscapes \texttt{labelIds} include additional class ids outside that range, so we normalize labels before computing loss.

\section*{Current Mapping Rule (Implemented)}
We use this rule in \texttt{convert\_label()}:
\begin{verbatim}
label[(label >= num_classes) & (label != ignore_index)] = ignore_index
\end{verbatim}
with:
\begin{itemize}
    \item \texttt{num\_classes = 20}
    \item \texttt{ignore\_index = 255}
\end{itemize}

This means:
\begin{enumerate}
    \item Labels \texttt{0..19} are kept as trainable classes.
    \item Labels \texttt{>= 20} are mapped to \texttt{255} and treated as ignored pixels.
\end{enumerate}

\section*{Why This Is Cleaner}
Compared to forcing unknown labels into a normal class, this approach avoids training the model with incorrect supervision for out-of-range categories. Unknown categories are simply excluded from optimization.

\section*{Loss Function Behavior}
Training uses:
\begin{verbatim}
criterion = nn.CrossEntropyLoss(ignore_index=255)
\end{verbatim}
So pixels labeled \texttt{255} do not contribute to gradients.

\section*{mIoU Behavior With Ignore Pixels}
Our mIoU computation first masks out ignored pixels (\texttt{target == 255}) before intersection/union is computed. Therefore:
\begin{itemize}
    \item ignored pixels do not affect per-class IoU,
    \item mIoU reflects only the supervised class range \texttt{0..19}.
\end{itemize}

\section*{Discussion Summary}
\begin{itemize}
    \item We still need \texttt{convert\_label()} because raw labels contain ids outside the model class range.
    \item We do not directly use every raw label id for training in this simplified setup.
    \item The new rule is more consistent: out-of-range labels are ignored, not reassigned to a trainable class.
    \item mIoU remains useful for comparing models in this pipeline because it is computed with the same ignore policy.
\end{itemize}

\section*{Training History and mIoU Summary (from \texttt{outputs/histroy\_and\_miou.txt})}
The training logs cover 8 models, each trained for 20 epochs. Across models, train/validation losses generally decrease over time with expected fluctuations, and checkpoints are saved whenever validation loss improves.

Final mIoU on 100 validation samples:
\begin{itemize}
    \item \texttt{vgg16\_dialated.pt}: 0.4261 (best mIoU)
    \item \texttt{vgg16.pt}: 0.4209
    \item \texttt{lraspp.pt}: 0.3846
    \item \texttt{lraspp\_dialated.pt}: 0.3658
    \item \texttt{segnet\_dialated.pt}: 0.3484
    \item \texttt{segnet.pt}: 0.3315
    \item \texttt{dialted.pt}: 0.2331
    \item \texttt{simple.pt}: 0.2047
\end{itemize}

Observed pattern: dilation improves Simple, VGG16, and SegNet variants, but does not improve LRASPP in this setup.

\section*{Inference Time and Parameter Summary (from \texttt{outputs/inference\_time\_and\_parameters.txt})}
Inference was measured on the original resolution (\texttt{1024x2048}) for 10 validation samples on CUDA.

\begin{center}
\small
\begin{tabular}{|l|r|r|r|}
\hline
Model & mIoU & Parameters & Time for 10 samples (s) \\
\hline
\texttt{simple.pt} & 0.2047 & 135,379 & 0.1969 \\
\texttt{dialted.pt} & 0.2331 & 135,379 & 0.1970 \\
\texttt{vgg16.pt} & 0.4209 & 8,367,379 & 1.0597 \\
\texttt{vgg16\_dialated.pt} & 0.4261 & 8,367,379 & 1.0652 \\
\texttt{segnet.pt} & 0.3315 & 468,403 & 0.5819 \\
\texttt{segnet\_dialated.pt} & 0.3484 & 468,403 & 0.6221 \\
\texttt{lraspp.pt} & 0.3846 & 3,221,198 & 0.1658 \\
\texttt{lraspp\_dialated.pt} & 0.3658 & 3,221,198 & 0.1643 \\
\hline
\end{tabular}
\end{center}

\section*{Why LRASPP Has the Best Inference Time}
The measured best inference time is from \texttt{lraspp\_dialated.pt} (0.1643s for 10 samples), followed by \texttt{lraspp.pt} (0.1658s). This is faster than \texttt{simple.pt}/\texttt{dialted.pt} (about 0.197s), even though LRASPP has many more parameters (3.22M vs 0.135M).

This happens because parameter count and runtime are not equivalent:
\begin{itemize}
    \item Runtime is dominated by actual compute and memory movement at each feature-map size, not only by total weights.
    \item LRASPP (MobileNetV3-based) relies heavily on efficient depthwise-separable operations and early downsampling, which reduces high-resolution computation.
    \item The Simple/Dialted models use regular dense convolutions on large spatial maps, which can cost more compute per pixel even with far fewer parameters.
    \item On GPU, optimized kernels for MobileNet-style operators can further reduce latency.
\end{itemize}

So in this project, LRASPP gives a strong speed/accuracy trade-off: much faster inference than Simple/Dialted while achieving substantially higher mIoU.

\end{document}
